Data cleaning Study 3

### 2024-05-31 exp_2/cleaning.Rmd
preregistration study 3
first write up studies 1 to 3

---
title: "Data cleaning Study 1"
output: 
  html_document: 
    keep_md: yes
---

@@

There are 6 failed attention checks.

@@

>>>>

---
title: "Data cleaning Study 2"
output: 
  html_document: 
    keep_md: yes
---

@@

There are 11 failed attention checks.

@@

## Justifications

### Extract justifications

```{r}
# questions <- c("electrons", "antibiotics", "continents", "sex", "lasers", "orbit", "diamonds", "speed", "salt", "water")
# 
# justifications <- exp2_wide %>% 
#   select(id, all_of(questions)) %>% 
#   pivot_longer(all_of(questions), 
#                names_to = "question", 
#                values_to = "answer") %>% 
#   drop_na(answer) %>% 
#   arrange(id)
# 
# write_csv(justifications, "data/justifications.csv")
```

### Clean justifications

Note that we had used the following coding scheme (where 4 would be the top-category for 5 and 6, i.e. 4 has never been coded):

1	- no justification provided / no clear justification	
2	- mistake, they actually accept the consensus	
3	- not convinced by the explanation provided	
4	- motivated rejection of the belief	
5	-- do to personal convictions	
6	-- for religious reasons

```{r}
justifications <- read_csv("data/justifications_coded.csv")

# Research questions
justifications <- justifications %>%
  mutate(category = case_when(category == 1 ~ "No justification",
                              category  == 2 ~  "Mistake",
                              category  == 3 ~ "Not convinced",
                              category  == 5 ~ "Personal convictions",
                              category  == 6 ~ "Religious Beliefs",
                              TRUE ~ NA_character_
  )
  )

write_csv(justifications, "data/justifications_clean.csv")
```

### 2024-05-31 exp_3/analysis.Rmd
first write up studies 1 to 3

---
title: "Analysis Study 1"
output:
  bookdown::pdf_document2: default
  bookdown::word_document2: default
  bookdown::html_document2:
      keep_md: yes
always_allow_html: true
bibliography: ../references.bib
---

@@

# read wide version data set
data_wide <- read_csv("data/cleaned_wide.csv")

# rename d wide
d <- data_wide
d_avg <- data_wide
```

# RQ1

**RQ1: What is the average science knowledge score (1)?**

```{r}
# as table
# d_avg %>% 
#   mutate(avg_knowledge = as.factor(avg_knowledge)) %>% 
#   group_by(avg_knowledge) %>% 
#   count() %>% 
#   ungroup() %>% 
#   mutate(share = n / sum(n)) %>% 
#   mutate_if(is.numeric, round, digits = 3)

# as plot
# Calculate the mean
mean_value <- mean(d_avg$avg_knowledge)

ggplot(d_avg, aes(x = avg_knowledge)) +
  geom_bar() + 
  geom_vline(xintercept = mean_value, color = "red", linetype = "dashed") +
  geom_text(aes(x = mean_value, y = 40, label = paste("Mean =", round(mean_value, 2))), 
            vjust = -0.5, hjust = 1.2) +
  scale_x_continuous(breaks = seq(0, 1, 0.1)) +
  labs(x = "Average science knowledge", 
       y = "N")
```

# RQ2

**RQ2: What is the average acceptance of the scientific consensus (2)?**

```{r}
# Calculate the mean
mean_value <- mean(d_avg$avg_acceptance)

ggplot(d_avg, aes(x = avg_acceptance)) +
  geom_bar() + 
  geom_vline(xintercept = mean_value, color = "red", linetype = "dashed") +
  geom_text(aes(x = mean_value, y = 60, label = paste("Mean =", round(mean_value, 2))), 
            vjust = -0.5, hjust = 1.2) +
  scale_x_continuous(breaks = seq(0, 1, 0.1)) +
  labs(x = "Average consensus acceptance", 
       y = "N")
```

# RQ3

**RQ3: What is the relationship between trust in science and, respectively, knowledge and acceptance?**

```{r}
plot_data <- d %>% 
  group_by(wgm_sciencegeneral, avg_knowledge) %>% 
  summarise(n = n())

# Plot as a tile plot
ggplot(plot_data, aes(x = avg_knowledge, y = wgm_sciencegeneral, fill = n)) +
  geom_tile() +  # Add black border around tiles for better visibility
  geom_jitter(inherit.aes = FALSE, data = d, aes(x = avg_knowledge, y = wgm_sciencegeneral),
              width = 0.03, height = 0.08) +  # Scatter points within each tile
    geom_smooth(method = "lm") +
  scale_fill_viridis_c(option = "plasma") +  # Use Viridis color scale (Plasma)
  scale_x_continuous(breaks = seq(0, 1, 0.1)) +
  scale_y_continuous(breaks = 1:5) +
  labs(x = "Average science knowledge", 
       y = "Trust in science", 
       fill = "N") +  # Update fill legend title
  theme_minimal() 
```

```{r}
plot_data <- d %>% 
  group_by(wgm_sciencegeneral, avg_acceptance) %>% 
  summarise(n = n())

# Plot as a tile plot
ggplot(plot_data, aes(x = avg_acceptance, y = wgm_sciencegeneral, fill = n)) +
  geom_tile() +  # Add black border around tiles for better visibility
  geom_jitter(inherit.aes = FALSE, data = d, aes(x = avg_acceptance, y = wgm_sciencegeneral),
              width = 0.03, height = 0.08) +  # Scatter points within each tile
  scale_fill_viridis_c(option = "plasma") +  # Use Viridis color scale (Plasma)
  scale_x_continuous(breaks = seq(0, 1, 0.1)) +
  scale_y_continuous(breaks = 1:5) +
  labs(x = "Average consensus Acceptance", 
       y = "Trust in science", 
       fill = "N") +  # Update fill legend title
  theme_minimal() 
```

## Difference between false and true answer acceptance

```{r}
# make a version with averages per participant, grouped by false/true ansers 
veracity_data_avg <- data_long %>% 
  # make numeric versions
  mutate(acceptance_num = ifelse(acceptance == "Yes", 1, 0)
  ) %>% 
  group_by(id, knowledge) %>% 
  # calculate by-participant averages
  summarize(
    n = n(),
    n_accepted = sum(acceptance_num),
    avg_acceptance  = sum(acceptance_num)/n(), 
    wgm_sciencegeneral = mean(wgm_sciencegeneral)
  ) %>% 
  mutate(knowledge = ifelse(knowledge == FALSE, "False answers", "Correct answers"))

# make data frame with average particpant data and wide format
# false_data <- left_join(false_data, false_data_avg, by = "id")

ggplot(veracity_data_avg, aes(x = avg_acceptance, y = wgm_sciencegeneral)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_x_continuous(breaks = seq(0, 1, 0.1)) +
  labs(x = "Average consensus acceptance", 
       y = "Average trust in science") +  # Update fill legend title
  theme_minimal() +
  facet_wrap(~knowledge)

```

# RQ4

**RQ4: What is the relationship between conspiracy thinking and, respectively, knowledge and acceptance?**

```{r}
# Plot as a tile plot
ggplot(d, aes(x = avg_knowledge, y = BCTI_avg)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_x_continuous(breaks = seq(0, 1, 0.1)) +
  labs(x = "Average science knowledge", 
       y = "Average conspiracy thinking") +  # Update fill legend title
  theme_minimal() 
```

```{r}
ggplot(d, aes(x = avg_acceptance, y = BCTI_avg)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_x_continuous(breaks = seq(0, 1, 0.1)) +
  labs(x = "Average consensus acceptance", 
       y = "Average conspiracy thinking") +  # Update fill legend title
  theme_minimal() 
```

## Difference between false and true answer acceptance

```{r}
# make a version with averages per participant, grouped by false/true ansers 
false_data_avg <- data_long %>% 
  # make numeric versions
  mutate(acceptance_num = ifelse(acceptance == "Yes", 1, 0)
  ) %>% 
  group_by(id, knowledge) %>% 
  # calculate by-participant averages
  summarize(
    n = n(),
    n_accepted = sum(acceptance_num),
    avg_acceptance  = sum(acceptance_num)/n(), 
    BCTI_avg = mean(BCTI_avg)
  ) %>% 
  mutate(knowledge = ifelse(knowledge == FALSE, "False answers", "Correct answers"))

# make data frame with average particpant data and wide format
# false_data <- left_join(false_data, false_data_avg, by = "id")

ggplot(false_data_avg, aes(x = avg_acceptance, y = BCTI_avg)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_x_continuous(breaks = seq(0, 1, 0.1)) +
  labs(x = "Average consensus acceptance", 
       y = "Average conspiracy thinking") +  # Update fill legend title
  theme_minimal() +
  facet_wrap(~knowledge)

```

Alternative plot

```{r}
plot_data <- data_long %>% 
  mutate(BCTI_avg = round(BCTI_avg, digits = 0)) %>% 
  group_by(knowledge, acceptance, BCTI_avg) %>% 
  count() %>% 
  group_by(knowledge, BCTI_avg) %>% 
  mutate (share = n/sum(n)) %>% 
  mutate_if(is.numeric, round, digits = 3) %>% 
  mutate(knowledge = ifelse(knowledge == FALSE, "False answers", "Correct answers"))

ggplot(plot_data, aes(x = BCTI_avg, y = share, fill = acceptance)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format(), breaks = seq(0, 1, 0.1)) +
  scale_x_continuous(breaks = seq(1, 9, 1)) +
  geom_text(aes(label = paste0(round(share * 100, 1), "%")),
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3) +
  scale_fill_viridis_d(option = "cividis") +
  labs(x = "Average conspiracy thinking", fill = "Acceptance \nof consensus") +
  theme_minimal() +
  facet_wrap(~knowledge) +
  theme(legend.position = "top")

ggplot(data_long, aes(x = ))
```

# Additional RQs

## Knowledge and Acceptance by science question

```{r}
# calculate share
plot_data <- data_long %>% 
  group_by(subject) %>% 
  summarize(share_true = sum(knowledge == TRUE, na.rm = TRUE)/n(),
            share_accepted = sum(acceptance == "Yes", na.rm = TRUE)/n()) 
```

```{r}
# plot knowledge
ggplot(plot_data, aes(x = reorder(subject, share_true), y = share_true)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent_format(), breaks = seq(0, 1, 0.1)) +
  labs(x = "Question", y = "Share of correct responses") +
  coord_flip()
```

```{r}
# plot acceptance
ggplot(plot_data, aes(x = reorder(subject, share_accepted), y = share_accepted)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent_format(), breaks = seq(0, 1, 0.1)) +
  labs(x = "Question", y = "Share of acceptance") +
  coord_flip()
```

## Open ended comments
```{r}
levels(as.factor(d$comments))
```

## Is the share of correct responses associated with the share of acceptance ?

```{r}
plot_data <- d %>% 
  group_by(avg_knowledge, avg_acceptance) %>% 
  summarise(n = n())

# Plot as a tile plot
ggplot(plot_data, aes(x = avg_knowledge, y = avg_acceptance, fill = n)) +
  geom_tile() +  # Add black border around tiles for better visibility
  geom_jitter(inherit.aes = FALSE, data = d, aes(x = avg_knowledge, y = avg_acceptance),
              width = 0.03, height = 0.04) +  # Scatter points within each tile
  scale_fill_viridis_c(option = "plasma") +  # Use Viridis color scale (Plasma)
  scale_x_continuous(breaks = seq(0, 1, 0.1)) +
  scale_y_continuous(breaks = seq(0, 1, 0.1)) +
  labs(x = "Average science knowledge", 
       y = "Average consensus acceptance", 
       fill = "N") +  # Update fill legend title
  theme_minimal() 
```

## What were the acceptance rates among true and false answers?

@@

ggplot(plot_data, aes(x = knowledge, y = share, fill = acceptance)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format(), breaks = seq(0, 1, 0.1)) +
  geom_text(aes(label = paste0(round(share * 100, 1), "%")),
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3) +
  scale_fill_viridis_d(option = "cividis") +
  labs(x = "Knowledge answer", fill = "Acceptance \nof consensus") +
  theme_minimal()
```

## To which extent did people believe the different conspiracy items ?

```{r}
BCTI_data <- data_long %>% 
  # bring to long format
  pivot_longer(cols = BCTI_apollo:BCTI_evolution, 
               names_to = "BCTI_item", values_to = "score") %>% 
  select(id, c(starts_with("BCTI"), score)) %>% 
  group_by(BCTI_item) %>% 
  summarize(avg_score = mean(score)) %>% 
  ungroup()

ggplot(BCTI_data, aes(x = reorder(BCTI_item, avg_score), y = avg_score)) +
    geom_pointrange(aes(ymin = 1, ymax = avg_score),
                  # Make the lines a little thicker and the dots a little bigger
                  fatten = 2, size = 1) +
  scale_y_continuous(limits = c(1, 9), breaks = seq(1, 9, 1)) +
  labs(x = "Conspiracy item", y = "Average score") +
  coord_cartesian(ylim = c(1, 9)) +
  coord_flip()
```

# Redo all analyses without tree question

>>>>

---
title: "Analysis Study 3"
output:
  bookdown::pdf_document2: default
  bookdown::word_document2: default
  bookdown::html_document2:
      keep_md: yes
always_allow_html: true
bibliography: ../references.bib
---

@@

# read wide version data set
data_wide <- read_csv("data/cleaned_wide.csv") 
```

@@

absolute_numbers <- plot_data %>% 
  group_by(knowledge) %>% 
  summarise(n = sum(n))

ggplot(plot_data, aes(x = knowledge, y = share, fill = acceptance)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format(), breaks = seq(0, 1, 0.1)) +
  geom_text(aes(label = paste0(round(share * 100, 1), "%")),
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3) +
  scale_fill_viridis_d(option = "cividis") +
  labs(x = "Knowledge answer", fill = "Acceptance \nof consensus") +
  theme_minimal() +
  geom_bracket(data = absolute_numbers,
               aes(xmin = as.numeric(factor(knowledge)) - 0.4, 
                   xmax = as.numeric(factor(knowledge)) + 0.4, 
                   y.position = 1.05, label =  paste0("n = ", n)),
               inherit.aes = FALSE,
               step.increase = 0)
```

### 2024-05-31 exp_3/cleaning.Rmd
first write up studies 1 to 3

## Import data
```{r import, message=FALSE}
d <- read_csv("data/test_study_3.csv")
names(d)
```

@@

@@

```{r}
d_long <- d_wide %>% 
  # bring to long format
  pivot_longer(cols = c(ends_with("know"), ends_with("accept"), ends_with("confirm")), 
               names_to = "subject_question", values_to = "answer") %>% 
  # separate conditions into CONVERGENCE_OUTCOME_STIMULUS
  separate_wider_delim(subject_question, "_", names = c("subject", "question")
                       ) %>%
  pivot_wider(names_from = question, values_from = answer) %>% 
# create better variable names
  rename(knowledge = know, 
         acceptance_initial = accept, 
         acceptance_confirmation = confirm) %>% 
  # make a final acceptance variable
  mutate(acceptance = ifelse(is.na(acceptance_confirmation), acceptance_initial, acceptance_confirmation))
```

@@

```{r}
# make a version with averages per participant
d_avg <- d_long %>% 
  # make numeric versions
  mutate(acceptance_num = ifelse(acceptance == "Yes", 1, 0), 
         knowledge_num = ifelse(knowledge == TRUE, 1, 0)
  ) %>% 
  group_by(id) %>% 
  # calculate by-participant averages
  summarize(avg_acceptance  = sum(acceptance_num)/n(), 
            avg_knowledge = sum(knowledge_num)/n()
  )

# make data frame with average particpant data and wide format
d_wide <- left_join(d_wide, d_avg, 
               by = "id")
```

>>>>

## Import data
```{r import, message=FALSE}
d <- read_csv("data/qualtrics.csv")
names(d)
```

@@

Code the reasons for why people accept the scientific consensus:

```{r}
d_wide <- d_wide %>%
  mutate(reason_agreement = case_when(
    grepl("Other:", reason_agreement) ~ "other",
    grepl("independently", reason_agreement) ~ "independent verification",
    grepl("trust scientists", reason_agreement) ~ "trust in scientists",
    TRUE ~ NA_character_  # This handles any NAs
  )) %>% 
  rename(other_reason_agreement = reason_agreement_3_TEXT)
```

@@

```{r}
d_long <- d_wide %>% 
  # bring to long format
  pivot_longer(cols = c(ends_with("know"), ends_with("accept"), ends_with("confirm")), 
               names_to = "subject_question", values_to = "answer") %>% 
  # separate subject (e.g. water) and question (e.g. acceptance)
  separate_wider_delim(subject_question, "_", names = c("subject", "question")
                       ) %>%
  pivot_wider(names_from = question, values_from = answer) %>% 
# create better variable names
  rename(knowledge = know, 
         acceptance_initial = accept, 
         acceptance_confirmation = confirm) %>% 
  # make a final acceptance variable
  mutate(acceptance = ifelse(is.na(acceptance_confirmation), acceptance_initial, acceptance_confirmation))
```

@@

```{r}
# make a version with averages per participant
d_avg <- d_long %>% 
  # make numeric versions
  mutate(acceptance_num = ifelse(acceptance == "Yes", 1, 0), 
         knowledge_num = ifelse(knowledge == TRUE, 1, 0), 
         acceptance_initial_num = ifelse(acceptance_initial == "Yes", 1, 0)
  ) %>% 
  group_by(id) %>% 
  # calculate by-participant averages
  summarize(avg_acceptance  = sum(acceptance_num)/n(), 
            avg_knowledge = sum(knowledge_num)/n(),
            avg_acceptance_initial  = sum(acceptance_initial_num)/n()
  )

# make data frame with average participant data and wide format
d_wide <- left_join(d_wide, d_avg, 
               by = "id")
```

## Check NAs

```{r}
d_wide %>% 
  summarize(across(c("wgm_scientists", "wgm_sciencegeneral", "pew", "reason_agreement", "BCTI_avg", "CMQ_avg", "SICBS", "avg_acceptance", "avg_knowledge"), 
                   ~sum(is.na(.x))
                   )
            )
```

It's technically not possible to have NA's for trust in science and conspiracy questions, because we used forced responses on them. Checking the qualtrics, we know that we accidentally randomized people into two of three outcome measures: trust in science, conspiracy or reasons for accepting consensus. This leaves us with reduced sample sizes for all analyses concerning these outcomes (XX for trust in science question, XX for conspiracy measures, XX for accepting consensus).

```{r}
d_wide %>% 
  summarize(across(c("wgm_scientists", "wgm_sciencegeneral", "pew", "reason_agreement", "BCTI_avg", "CMQ_avg", "SICBS"), 
                   ~sum(!is.na(.x))
                   )
            ) %>% 
  pivot_longer(everything(), 
               names_to = "outcome", 
               values_to = "n_valid")
```

## Clean justifications

```{r}
justifications <- read_csv("data/justifications_coded.csv")

# Research questions
justifications <- justifications %>%
  mutate(category = case_when(final_coding == 1 ~ "No justification",
                              final_coding == 2 ~  "Mistake",
                              final_coding == 3 ~ "Not convinced",
                              final_coding == 4 ~ "Personal convictions",
                              final_coding == 5 ~ "Religious Beliefs",
                              TRUE ~ NA_character_
                              )
         )

write_csv(justifications, "data/justifications_clean.csv")
```
